---
title: "Pass 1B Lung Phospho analysis for MoTrPAC."
author: "Michael Nestor"
---



```{r setup, echo=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
# knitr::opts_chunk$set(echo=T, message=F, warning=F, fig.align='center', out.width='10cm')
```

# Loading crosstab

``` {r expression_sets}
library(Biobase)
library(readxl)
library(dplyr)
library(BiocParallel)
library(MSnbase)
library(vp.misc)
ratio_results <-  read.table("MoTrPAC_PASS1B_Lung_Phospho_results_ratio.txt",
                             header=T)
rownames(ratio_results) <- ratio_results$ptm_id
ratio_results <- ratio_results %>%
  dplyr::select(-ptm_id, -protein_id, -gene_symbol, -entrez_id, -confident_score, -confident_site)
colnames(ratio_results) <- sub("X", "", colnames(ratio_results))

m <- MSnSet(as.matrix(ratio_results))
```

# Attaching metadata

```{r}
phenoData <- read_excel("./../PASS1B_T66_lung_metadata.xlsx")

phenoData$plex <- as.factor(phenoData$plex)
phenoData$vialLabel <- as.factor(phenoData$vialLabel)
phenoData$training <- factor(phenoData$training, levels=c("control","1wk","2wk","4wk","8wk"))

phenoData <- as.data.frame(phenoData)
rownames(phenoData) <- paste0(phenoData$vialLabel)

phenoData <- phenoData[sampleNames(m),]
pData(m) <- phenoData
validObject(m)
```

# Normalization

(Partial) global normalization coefficients are used.

```{r normalize}
global_ratio_results <- read.table("./../global/MoTrPAC_PASS1B_Lung_Global_results_ratio.txt", header=T)

rownames(global_ratio_results) <- global_ratio_results$protein_id
global_ratio_results <- global_ratio_results %>%
  dplyr::select(-protein_id, -gene_symbol, -entrez_id)
colnames(global_ratio_results) <- sub("X", "", colnames(global_ratio_results))
exprs_global <- as.matrix(global_ratio_results)

global_norm_coefficients <- apply(exprs_global, 2, median, na.rm=T)
exprs(m) <- sweep(exprs(m), 2, global_norm_coefficients, "-")
```

```{r saving}
saveRDS(m, "msnset.Rds")
```

# Batch correction

This section explores batch effect using PCA.

```{r pca_plot1}
library(ggplot2)
library(ggpubr)
library(vp.misc)
plot_pca_v3(m, label="channel", phenotype="plex") +
  ggtitle("PCA of PASS 1B Lung (phospho), grouped by plex")
```

There appears to be moderate batch effect. Let's remove it using `vp.misc::remove_batch_effect`.

```{r remove_batch_effect}
m <- vp.misc::remove_batch_effect(m, "plex")
```

Now check the PCA again:

```{r pca_plot2}
plot_pca_v3(m, label="channel", phenotype="plex") +
  ggtitle("PCA of PASS 1B Lung (phospho), grouped by plex after batch correction")
```

Looks good!

# Sex and training time

First let's take a look at PCA by sex.

```{r pca_plot3}
plot_pca_v3(m, label="channel", phenotype="sex") +
  ggtitle("PCA of PASS 1B Lung (phospho), grouped by sex")
```

It appears significant. Lets look at differences in training time next.

```{r pca_plot4}
plot_pca_v3(m, label="channel", phenotype="training") +
  ggtitle("PCA of PASS 1B Lung (phospho), grouped by training")
```

It looks like the sex effect is dominating, so let's correct for sex first before making the PCA plot.

```{r pca_plot5}
m_sex_corrected <- vp.misc::remove_batch_effect(m, "sex")
plot_pca_v3(m_sex_corrected, label="channel", phenotype="training") +
  ggtitle("PCA of PASS 1B Lung (phospho), grouped by training and corrected for sex")
```

It is not really clear how big the effect of training is. Limma testing should provide a more analytical approach.

## Limma testing

```{r compute_limma}
limma <- vp.misc::limma_gen(m, "~sex + training", "sex")

ggplot(data=limma, aes(x=P.Value)) + geom_histogram(aes(y = ..density..)) +
  xlab("Unadjusted p-value") +
  ggtitle("Pass 1B Lung (phospho) p-value histogram for sex")
```

```{r top_significant_phosphosites}
limma %>%
  arrange(adj.P.Val) %>%
  dplyr::select(logFC, adj.P.Val) %>%
  head(20)
```

```{r compute_limma2}
limma <- vp.misc::limma_gen(m, "~sex + training", "training")

ggplot(data=limma, aes(x=P.Value)) + geom_histogram(aes(y = ..density..)) +
  xlab("Unadjusted p-value") +
  ggtitle("Pass 1B Lung (phospho) p-value histogram for training")
```

```{r cleanup, echo=FALSE}
unlink(".Rcache", recursive=TRUE)
```